---
title: "Comparative Analysis of the Homeless Veteran and Non-Veteran Demographic Profiles and Disabilitiesat Urban Ministries of Durham"
author: "Meagan Foster"
date: "December 1, 2019"
output: html_document
---

  ```{r}
source("master_df.R")

```

## Objectives

.Homeless veterans in the US

.Compare race, gender, and age for non-veteran and veteran clients at UMD

.Compare prevalence of disabilities for non-veteran and veteran clients 

.Summary

See ReadMe.md for more information on data sources and project structure.


##UMD Data Summary

  ```{r}
#Veteran Status by Gender
veteranstatus_gender_plot
```


  ```{r}
#Veteran Status by Race
veteranstatus_race_plot
```


  ```{r}
#Veteran Status by Age
veteranstatus_age_plot
```


  ```{r}
#Veteran Status by Age
disability_plot
```



  ```{r}
#Veteran Status by Age
disability_master_plot
```



  ```{r}
#Veterans with Disabilities
disability_veteran_plot
```


##Conclusion

Conduct a better count nationally. HUD's count should:

.Be nationally coordinated with a more consistent and morerigorous methodology. This and requires appropriate fundinglevels in order to get more useful data.

.Include estimation techniques designed and overseen byexperts in order to quantify the number of homeless individualsthat were missed during the count.

.Include all people experiencing homelessness, includingindividuals that are institutionalized in hospitals and jails orprisons
